### 2016-03-07 03-design.tex
Move HPC Cloud descr to beginning of 04

%\looseness=-1
Modern network technology allows for Remote Direct Memory Access (RDMA)
through the network cards (NICs) with negligible
overhead~\cite{Hamada_infinibandtrade,Beck:2011:PER:2043535.2043537}. The use
of RDMA enables communicating peers to avoid the high latency associated with
interrupting a processor and waiting for it to act upon a request. Instead, a
process would register a region of memory that can be directly accessed by the
NIC. However, RDMA only provides primitive operations and limited abstractions.
This has opened up the development of
efficient remote memory services, often in the form of distributed key-value
(DKV) stores.

The distributed algorithm implementation stores $\pi$ in a DKV store in the
collective memory of the
cluster, where $\pi[i]+\sum\phi[i]$ is the value for key $i$.
%
We decided not
to use any of the existing implementations of RDMA DKV stores, and instead
build our own DKV store on top of the Infiniband ib-verbs~API.
%
This stems from the fact that our use case is unique in a number
of important aspects. First, the number of keys and values to be stored are
constant throughout the execution of the algorithm. For instance, no records
are dynamically inserted or deleted from the DKV after their initial insertion.
However, the value associated with each key can be updated. Additionally, all
keys are of the same type and represent a unique vertex in the graph.
Therefore, keys can be represented as integer identifiers. Moreover, each key
points to a vector of $K+1$ floats as its value.
%
This data representation can be exploited to simplify the management of
read and write requests. Similarly, the keys can be deterministically
partitioned across the cluster using a fast hash function.

@@

%\looseness=-1
These updates are used by \textit{update\_pi} (line 7 in
Algorithm~\ref{algorithm}) locally, besides the value of
$\pi/\phi$ for the mini-batch vertices. The update to $\pi$ requires
$M \times K$ operations, and it is done in parallel over mini-batch vertices. Because of
memory consistency, this stage awaits completion of \textit{update\_phi} with
an MPI barrier. After the calculation, the updated values of $\pi;\sum\phi$
are written to the DKV store.

>>>>

%\looseness=-1
Modern network technology allows for Remote Direct Memory Access (RDMA)
through the network cards (NICs) with negligible
overhead~\cite{Hamada_infinibandtrade,Beck:2011:PER:2043535.2043537}. The use
of RDMA enables communicating peers to avoid the high latency for
interrupting a processor and waiting for it to act upon a request. Instead, a
process registers a region of memory that can be directly accessed by the
NIC. % RDMA only provides primitive operations and limited abstractions.
This has opened up the development of
efficient remote memory services, often in the form of distributed key-value
(DKV) stores.

The distributed algorithm implementation stores $\pi$ in a DKV store in the
collective memory of the
cluster, where $\pi[i]+\sum\phi[i]$ is the value for key $i$.
%
We decided not
to use any of the existing implementations of RDMA DKV stores, and instead
build our own DKV store on top of the Infiniband ib-verbs~API.
%
This is motivated by the fact that our use case is unique in a number
of important aspects. First, the number of keys and values to be stored are
constant throughout the execution of the algorithm. For instance, no records
are dynamically inserted or deleted from the DKV after their initial insertion,
although the value associated with each key can be updated. Additionally, all
keys are of the same type and represent a unique vertex in the graph.
Therefore, keys can be represented as integer identifiers. Moreover, each key
points to a vector of $K+1$ floats as its value.
%
This data representation can be exploited to simplify the management of
read and write requests. Similarly, the keys can be deterministically
partitioned across the cluster using a fast hash function.

@@

%\looseness=-1
These updates are used by \textit{update\_pi} (line 7 in
Algorithm~\ref{algorithm}) locally, besides the value of
$\pi/\phi$ for the mini-batch vertices. The update to $\pi$ requires
$M \times K$ operations, and it is done in parallel over mini-batch vertices. Because of
memory consistency, this stage awaits completion of \textit{update\_phi} with
an MPI barrier. After the calculation, the updated values of $\pi+\sum\phi$
are written to the DKV store.

### 2016-03-07 04-evaluation.tex
Move HPC Cloud descr to beginning of 04

Empirical results were obtained by performing experiments on the VU and
Leiden University
DAS5 clusters which consist of 68 and 24 compute nodes respectively. Each
compute node is equipped with a dual 8-core Intel Xeon E5-2630v3 CPU clocked
at 2.40GHz, 64GB of memory and 8TB of storage. Moreover, the compute nodes of
each site are interconnected by
FDR~InfiniBand. The MPI implementation used is MVAPICH2, which is optimally
tuned for Infiniband~\cite{mvapich}. All of the experiments reported in this section utilized
publicly available graphs from the Stanford Large Network Dataset
Collection~(SNAP) \cite{snapnets}. Table~\ref{tab-datasets} lists the
collection of data sets used.

@@

A study of a system's weak scaling aids in the assessment of the communication
and synchronization overheads associated with the management of large clusters.
Similarly, it can expose complex issues that hinder performance at scale such
as load imbalance. To fairly evaluate the algorithm's weak scaling behavior we
conducted several executions varying the cluster size and the number of latent
communities proportionally. The number of communities is 192 times the number
of worker nodes in all runs.
This methodology ensures that each compute node
performs an approximately constant amount of computational work across all
configurations. However, the number and size of messages exchanged between the
nodes would vary significantly. Figure~\ref{fig-weak-scaling} presents the
average execution
time per algorithm iteration across different cluster sizes.
This figure shows that the system's overall
overhead is minimal. Additionally, the experimental results show that the
implementation is capable of achieving good speedups provided the input problem
is large enough for the given cluster size.

@@

%################################################
\subsection{Horizontal vs. Vertical Scalability}
\begin{figure}[t] % [htb]
  \centering
  \epsfig{file=plots/hpc-cloud.eps, width=\columnwidth}
  \caption{Performance comparison between the distributed implementation
  running on DAS5 and the multi-threaded solution on machine with 40 cores and
  1TB of RAM.}
  \label{fig-scale-up}
\end{figure}
One of the main drawbacks of designing a distributed solution for a given
algorithm is the inherent complexity of communication and synchronization. A
considerably simpler approach would be developing a multi-threaded version and
running it on a machine with abundant memory and CPU cores. In such a context,
access to all of the algorithm's state would be an order of magnitude faster
than RDMA. Additionally, the overhead associated with synchronizing threads is
negligible compared to using MPI primitives. To evaluate the efficacy of both
approaches we utilized SURFsara's HPC Cloud system to instantiate a virtual
machine with 40 Intel Xeon E7-4850 cores clocked at 2.00GHz cores and 1TB of
memory. The physical machine underlying
the HPC Cloud system contains 40~CPU cores and does not oversubscribe
resources. Therefore, by provisioning all 40~cores we ensured that there is no
resource contention from other users of the system. Figure~\ref{fig-scale-up}
reports the execution time per algorithm iteration for two experimental setups.
%
First, Figure~\ref{fig-scale-up}-a shows the performance of executing the
algorithm on the HPC Cloud
system with 40 and 16 cores compared to a single DAS5 node with 16~cores. This
test uses the com-DBLP data set to enable the use of a large number of
communities without running out of memory. It is clear from the results that
the performance can benefit from the additional cores provided by HPC Cloud system.
%
Further, Figure~\ref{fig-scale-up}-b tests the performance of the HPC Cloud
system compared to 64 nodes of the DAS5
cluster using the com-Friendster data set. Clearly, the parallel and
distributed implementation vastly
outperforms the single-node multi-threaded solution. Moreover, the trajectory
of both curves shows a widening gap between them suggesting that the relative
performance difference will increase for larger $K$.
%
In conclusion, the overhead of network communication in the distributed version
is more than compensated by the increasing compute power compared to
a single-node implementation.

@@

The previous sections focused on the computational performance of the
distributed implementation. However, the algorithm's throughput does not
necessarily indicate how fast it can converge to a solution. More specifically, it
remains unclear how many iterations are needed for the algorithm to reach a
stable state and terminate. Therefore, we now shift our focus to the
convergence time to assess the system's utility. Figure~\ref{fig-ppx} presents
the convergence time of the data sets listed in Table~\ref{tab-datasets}.
%
The results in sub-figures a, b and c were obtained by using 65 compute nodes
of DAS5. In Figure~\ref{fig-ppx}-a, the number of communities was set to 12K
which fully occupied the aggregate memory resources of all 64 worker nodes since
com-Friendster has roughly 65 million vertices. In this case, the algorithm
reached a stable state after 3-4 hours. Next, Figures~\ref{fig-ppx}-b and
\ref{fig-ppx}-c present the convergence of com-LiveJournal and com-Orkut
respectively. As the number of vertices in these data sets is an order of
magnitude smaller than com-Friendster, we could use a larger number of
communities to fill up the collective memory. Naturally, the convergence time was extended as the complexity of
the algorithm increases dramatically with larger K. However, the system was
capable of reporting results for both in around 40 hours.
%
Figures~\ref{fig-ppx}-d, \ref{fig-ppx}-e and \ref{fig-ppx}-f were all
configured to use the number of ground-truth communities associated with
their respective data sets, com-Youtube, com-DBLP and com-Amazon. Since this
yields a much smaller storage requirement,
these experiments were conducted on 14, 24 and 24 cluster nodes respectively.

>>>>

Empirical results for the distributed implementation were obtained from experiments on the VU and
Leiden University
DAS5 clusters which consist of 68 and 24 compute nodes respectively. Each
compute node is equipped with a dual 8-core Intel Xeon E5-2630v3 CPU clocked
at 2.40GHz, 64GB of memory and 8TB of storage. Moreover, the compute nodes of
each site are interconnected by
FDR~InfiniBand. The MPI implementation used is OpenMPI, which supports
user-space Infiniband~\cite{gabriel04:_open_mpi}.

We compared the distributed performance against a shared-memory virtual machine
at SURFsara's HPC Cloud system with 40~cores and 1TB of memory; its underlying
physical system contains 40 Intel Xeon E7-4850 cores clocked at 2.00GHz, and
it does not oversubscribe resources.

All of the experiments reported in this section utilized
publicly available graphs from the Stanford Large Network Dataset
Collection~(SNAP) \cite{snapnets}. Table~\ref{tab-datasets} lists the
collection of data sets used.

@@

A study of a system's weak scaling aids in the assessment of the communication
and synchronization overheads associated with the management of large clusters.
Similarly, it can expose complex issues that hinder performance at scale such
as load imbalance. To fairly evaluate the algorithm's weak scaling behavior we
conducted several executions varying the cluster size and the number of latent
communities proportionally. The number of communities is 192 times the number
of worker nodes in all runs.
This methodology ensures that each compute node
performs an approximately constant amount of computational work across all
configurations. However, the number and size of messages exchanged between the
nodes varies significantly. Figure~\ref{fig-weak-scaling} presents the
average execution
time per algorithm iteration for different cluster sizes.
This figure shows that the system's overall
overhead is minimal. Additionally, the experimental results show that the
implementation is capable of achieving good speedups provided the input problem
is large enough for the given cluster size.

@@

%################################################
\subsection{Horizontal vs. Vertical Scalability}
\begin{figure}[t] % [htb]
  \centering
  \epsfig{file=plots/hpc-cloud.eps, width=\columnwidth}
  \caption{Performance comparison between the distributed implementation
  running on DAS5 and the multi-threaded solution on machine with 40 cores and
  1TB of RAM.}
  \label{fig-scale-up}
\end{figure}
One of the main drawbacks of designing a distributed solution for a given
algorithm is the inherent complexity of communication and synchronization. A
considerably simpler approach is developing a multi-threaded version and
running it on a machine with abundant memory and CPU cores. In such a context,
access to all of the algorithm's state would be an order of magnitude faster
than RDMA. Additionally, the overhead associated with synchronizing threads is
negligible compared to using MPI primitives. To evaluate the efficacy of both
approaches we used the virtual
machine with 40~cores and 1TB of
memory at SURFSara's HPC Cloud system. By provisioning all 40~cores
we ensured that there is no
resource contention from other users. Figure~\ref{fig-scale-up}
reports the execution time per algorithm iteration for two experimental setups.
%
First, Figure~\ref{fig-scale-up}-a shows the performance of executing the
algorithm on the HPC Cloud
system with 40 and 16 cores compared to a single DAS5 node with 16~cores. This
test uses the com-DBLP data set to enable the use of a large number of
communities without running out of memory. It is clear from the results that
the performance can benefit from the additional cores provided by HPC Cloud system.
%
Further, Figure~\ref{fig-scale-up}-b tests the performance of the HPC Cloud
system compared to 64 nodes of the DAS5
cluster using the com-Friendster data set. Clearly, the parallel and
distributed implementation vastly
outperforms the single-node multi-threaded solution. Moreover, the trajectory
of both curves shows a widening gap between them suggesting that the relative
performance difference will increase for larger $K$.
%
In conclusion, the overhead of network communication in the distributed version
is more than compensated by the increasing compute power compared to
a single-node implementation.

@@

The previous sections focused on the computational performance of the
distributed implementation. However, the algorithm's throughput does not
necessarily indicate how fast it can converge to a solution. More specifically, it
remains unclear how many iterations are needed for the algorithm to reach a
stable state and terminate. Therefore, we now shift our focus to the
convergence time to assess the system's utility. Figure~\ref{fig-ppx} presents
the convergence time of the data sets listed in Table~\ref{tab-datasets}.
%
The results in sub-figures a, b and c were obtained by using 65 compute nodes
of DAS5. In Figure~\ref{fig-ppx}-a, the number of communities was set to 12K
which fully occupied the aggregate memory resources of all 64 worker nodes since
com-Friendster has roughly 65 million vertices. In this case, the algorithm
reached a stable state after 8-10 hours. Next, Figures~\ref{fig-ppx}-b and
\ref{fig-ppx}-c present the convergence of com-LiveJournal and com-Orkut
respectively. As the number of vertices in these data sets is an order of
magnitude smaller than com-Friendster, we could use a larger number of
communities to fill up the collective memory. Naturally, the convergence time was extended as the complexity of
the algorithm increases dramatically with larger K. However, the system was
capable of reporting results for both in around 60 hours.
%
Figures~\ref{fig-ppx}-d, \ref{fig-ppx}-e and \ref{fig-ppx}-f were all
configured to use the number of ground-truth communities associated with
their respective data sets, com-Youtube, com-DBLP and com-Amazon. Since this
yields a much smaller storage requirement,
these experiments were conducted on 13, 21 and 21 cluster nodes respectively.
